---
title: "Class 12- Homework"
author: "Renee Zuhars (PID: A17329856)"
format: pdf
toc: TRUE
---

# Section 4: Population Scale Analysis

### Question 13: 
> Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}

data <- "rs8067378_ENSG00000172057.6.txt"

df_data <- read.table(data) # reading the downloaded data into R as a table

head(df_data) # Viewing the dataset to see what the row/column names are

nrow(df_data) # The number of rows in this dataset indicates how many samples there are. 

table(df_data$geno) # Seeing the table for each genotype will indicate how much of the sample size is represented by each genotype. 

library(dplyr) # Using tidyverse to sort first by genotype, and then by expression level- showing the mean levels of expression per genotype

df_data %>%
  group_by(geno) %>%
  summarise(median_exp = median(exp))
```
There are 462 total samples. Of these samples, 108 represent an A/A genotype, 233 represent an A/G genotype, and 121 represent a G/G genotype. 

The median levels of expression are as follows:
A/A: 31.24847	
A/G: 25.06486	
G/G: 20.07363	


### Question 14:
> Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

# Using ggplot to create a box plot for the downloaded data.frame

ggplot(df_data) +
  aes(x = geno, y = exp, fill = geno) +
  geom_boxplot(notch = TRUE)

# Below recommended by chatGPT: a quick statistical test to see if the genotype at rs8067378 correlates to expression of ORMDL3

# Results are significant if the p value < 0.05 - indicated by this test as * (<0.05), ** (<0.01), or *** (<0.001)

anova_result <- aov(exp ~ geno, data = df_data)
summary(anova_result)

```
From the boxplot above, it is seen that A/A is expressed at significantly higher levels than G/G. We may be able to infer from this that G/G is the recessive and more rare genotype to have. We could also infer that an A/A genotype is more advantageous to have, as natural selection has selected for it in a larger population of individuals. 

In accordance to a suggestion from ChatGPT, I ran a quick statistical test. Because the test returned a p-value of less than 0.001 (which is less than the usual threshold of <0.05), it can be assumed that having a different genotype at this location does affect expression of the gene ORMDL3. 






